The language of generalization: Probability, vagueness, and interaction

### 2018-03-18 paper/comgen-paper.Rmd
edits in model section

Generalizations are used to make predictions about events or properties of instances events that an agent has yet to experience [@HumeTHN].
People readily predict that the next dog will have four legs, drinking another cup of coffee will cause jitters, and a new day will find people that we know doing what they habitually do. 
In each case, we assign a specific exemplar $x$ to a category $k$, and make a prediction that it will have feature $f$.
Probability is the universal currency of belief and this prediction can be described by a conditional probability: $P(f \mid k) = P(x \in f \mid x \in k)$, the probability that $x$ will have $f$ given that it is in $k$ (formally, an instance of $k$ will be in the set of things that have $f$), which we will refer to as the *prevalence* and provide the notational simplification $p = P(f \mid k)$.
Our *prevalence* $p$ is a *prevalence in the mind*, a latent sense that a future instance of a category would have a particular property; others might call this projectibility [@Goodman1955].
The targets of our predictions can vary widely: They may be objects (a dog), events (a coffee drink), or more fine-grained types (a person on a particular day).
The properties also vary (e.g., having fur, causing jitters, going to the gym).
Yet the mathematical description of the inductive belief is always given by a probability $P(f \mid k)$.

<!-- a useful representation  -->
<!-- How do we come to know things?  -->
<!-- How does one acquire inductive beliefs? -->
Probability is a useful representation for human generalization from observations [@Shepard1987].
If you *observe* wugs (a novel category) that have two legs, you might infer that all wugs have two legs.
But not all properties have such strong projectibility (prevalence).
Seeing a wug with broken wings tells you comparatively less about other wugs having broken wings [@Nisbett1983].
Abstract, potentially domain-specific, beliefs about the projectibility of different properties can be represented by a probability distribution over the prevalence $P(p)$ [@Kemp2008].
Bayes Rule provides the mathematically correct way to update one's prior beliefs with some obesrvational data $d$: $P(p \mid d) \propto P(d \mid p) \cdot P(p)$, assuming some generative process that resulted in the observed data---the likelihood $P(d \mid p)$ [@Tenenbaum2001].

@@

The language of generalization is easy to spot when a property, that could project onto an individual, is predicated of a category (e.g., "Dogs have four legs"; *has four legs* could apply to an individual as in "This dog has four legs").
In the semantics literature on generics, bare plural sentences of this kind are sometimes described as *characterizing sentences* in contrast with *kind-denoting* sentences where the property can only meaningfully apply to the category as a whole (e.g., "Dinosaurs are extinct"; *extinct* cannot apply to an individual dinosaur).
<!-- Dinosaurs are exctinct ~ John is a smoker? -->
Generalization can also manifest when describing an instances of an individual [*habitual language*; e.g., "John smokes"; @Carlson1977, @Carlson2005].
Verbs like *causes* or *makes* also seem to convey generalization, in this case, about an instance of an actual causal event (*causal language*; e.g., "Staring at the sun makes you go blind"). 
Psychologists, linguists, and philosophers have long studied the language of generalization, as it appears very simple (e.g., syntactically) yet its meaning is difficult to formalize.^[
  We further distinguish the problem of formalizing a *meaning* for the language of generalization as opposed to the problem of *identifying* generalizations. 
  Syntax alone is neither necessary nor sufficient for a listener to know that the sentence conveys a generalization  (e.g., indefinite singulars can encode generalizations: "A dog has four legs" and bare plurals may not: "Dogs are on my front lawn").
  Our analysis thus begins once a sentence has been disambiguated as conveying a generalization. 
]

If generalization from observations can be described by a probability $p$, it is natural to posit that same construct will be at the heart of a semantic theory of the language of generalization.
The semantic tools of *truth values* provide a way to update a state of knowledge by relating the semantic quantity (probability) to Boolean values [@Montague1973].
The simplest way to relate a scalar quantity like probability is via a *threshold semantics*: The utterance is true if the relevant scalar value is above a threshold.
Thus, the simplest semantics for a generalization is a threshold on the prevalence: $\denote{generalization}(p, \theta) = \{p >\theta\}$.

This formalism builds on a tradition in formal semantics to try to precisely define the conditions under which a sentence is true or false [@Montague1973].
We go beyond previous semantic models of generics [e.g., @Cohen1999] by embedding this semantic hypothesis in a cognitive model in which relevant contextual knowledge can be formalized.
By doing so, we thus provide an account of how context influences generic understanding and are able to make precise, quantitative predictions about human judgments.

@@

We formalize this underspecified (or, *vague*) semantics in a Bayesian cognitive model, where the threshold $\theta$ is sampled from a uniform prior distribution over possible thresholds $P(\theta)$ [@Lassiter2013;   @Lassiter2015; cf., @Qing2014].^[
  Following @Lassiter2015, we defined $P(\theta)$ to be a uniform distribution over the support of $P(p)$.
]
The model also has prior knowledge about the property $P(p)$ (described in more detail below) and literal interpretation is given by updating these prior beliefs according the truth-functional meaning of utterance $\delta_{\denote{u}(p, \theta)}$:

@@

Given the above model of literal interpretation, how do we decide when a statement conveying a generalization is intuitively true or not?
We consider the notion of felicity or endorsement from the information-theoretic perspective: Would a generalization convey the appropriate information to the naive interpreter? 
We formalize this by considering how well the generalization would convey the prevalence $p$ to the $Lit$ model (Eq. \ref{eq:L0}).^[
  A more general version of this model can relax the assumption that the endorsement model has access to a specific prevalence $p$ that it wants to communicate.
  Rather, the endorser may have a distribution over $p$ for the category $k$, corresponding to the endorser's uncertain beliefs about the prevalence. 
  In this situation, we would define the endorsement model decision to be with respect to the expected value of the informativity, which integrates over the endorsement model's belief distribution:
  $S(u \mid k) \propto \exp{(\lambda \cdot {\mathbb E}_{p\sim P_{k}} \ln{ \int_{\theta} Lit(p, \theta \mid u)} \diff \theta )}$
  For the empirical case studies described below, these two versions of the model make almost identical predictions.
]

\begin{equation} 
S(u \mid p) \propto \exp{(\lambda \cdot \ln{ \int_{\theta} Lit(p, \theta \mid u)  \diff \theta } )}
\label{eq:S1}
\end{equation}

The endorsement model simulates what the interpretation model would believe upon hearing an utterance $u$: $Lit(p, \theta \mid u)$.
$S$ does not know $\theta$ and integrates over the interpreter's posterior beliefs about $\theta$.
Following Bayesian decision theory, we assume the decision to endorse is approximately rational, with degree of rationality given by $\lambda$.
The proportionality in Eq. \ref{eq:S1} implies normalization over a set of alternative utterances. 
We interpret the endorsement task (e.g., "true" vs "false"; "agree" vs. "disagree") as supplying two alternatives: *endorse* or *not*, where *not* we take to be a silent utterance that conveys no information content [@Franke2014cogsci; @Degen2014]: $\denote{null}(x, \theta) = \text{True}$.^[
  This "null" alternative to the generalization can be realized in at least two other ways: the negation of the generalization (e.g., "It is not the case that Robins lay eggs") or the negative generalization (i.e., "Robins do not lay eggs").
  The latter is different because it creates a new threshold (analogous to how "short" and "tall" are modeled with different thresholds).
  All results reported are similar for these two alternatives, and we use the alternative of the "silent" utterance for simplicity.
]

Our endorsement model (Eq. \ref{eq:S1}) provides a mapping from the prevalence in the category being referred to (i.e., how many robins lay eggs)---the *referent prevalence* $p$---to an endorsement probability for the corresponding generalization (e.g., "Robins lay eggs"): $S(u = u_{gen} \mid p)$. 
The model assumes a threshold semantics based on probability (prevalence) but does not require specifying the threshold *a priori* (note that there is no $\theta$ in the left-hand side of Eq. \ref{eq:S1}).
This model can be seen as a special case of a Rational Speech Act model [@Frank2012; @Goodman2016] with only a single level of recursion and the space of possible utterances maximally constrained; the endorsement model is a "speaker", while the interpretation model is a "listener".

@@

The uncertain, truth-conditional threshold $\theta$ is resolved by the literal interpretation model $Lit$ (Eq. \ref{eq:L0}) by integrating the model's prior beliefs about the feature in question---the *prevalence prior* $P(p)$---with the underspecified semantics.
Because $P(\theta)$ is assumed to be a uniform distribution across all contexts, a generalization could in principle be felicitous with any referent-prevalence. 
However, the interpretation model's prior beliefs $P(p)$ must be considered before the endorsement model can say whether or not would it want to update those beliefs by endorsing the statement.

@@

Beliefs about properties result from internal, conceptual models of the world. 
We do not take up the task of formalizing particular conceptual models, but rather merely try to measure their implications on beliefs about prevalence. 
Without appealing to domain specific knowledge, one can construct a prior distributions over prevalence by considering the prevalence of the feature among alternative categories.
For example, many different kinds of animals *do not bark* (more precisely: have close to 0 exemplars who bark); thus, the prevalence prior for the feature *barks* should have substantial probability mass near 0, because one would expect a randomly-sampled category to have 0 or close to 0 exemplars that bark.
Furthermore, among the categories that do have exemplars which bark (e.g., *dogs*, *seals*, ...), the probability is quite high that an exemplar will bark; thus, the corresponding prevalence prior would have some probability mass at relatively high prevalences, potentially with some uncertainty (Figure\ \@ref(fig:simulations); column 1).
Abstract beliefs between different kinds of categories could manifest in these prevalence priors in the form of multi-modal distributions; for example, the prevalence prior for *lays eggs* should similarly have substantial probability mass near 0 (since many animal categories do not have egg-layers) but should have a secondary component peaked around 0.5, because among the animals with egg-layers, only the female members of the category have the property (Figure\ \@ref(fig:simulations); column 4).

@@

```{r uniformPriorSimulation, fig.cap = "Computational model behavior. A: Interpretation model posteriors assuming different fixed thresholds. High thresholds rule out more world-states (prevalence levels), leaving fewer world-states among which to distribute the full probability mass. B: Full interpretation model marginalizes over possible thresholds to return a posterior distribution that favors higher prevalence levels in a graded manner. C: Endorsement model predicts higher rates of endorsemens as prevalence levels increase.", fig.width = 13, fig.asp = 0.3, out.width = "100%", fig.align = "center", cache = F}

@@

These simulations show how the model predicted endorsement depends upon the background knowledge of the interpreter about the property $P(p)$ as well as the referent-prevalence of the endorser $p$ communicated. 
We chose schematic values for each of these, and saw that the endorsement model recovers standard intuitions about the truth or falsity of classic examples from the generics literature.
In what follows, we test this theory for a wide range of generalizations, including generalizations of different types (categories, events, and causes).
We do this by both measuring and manipulating background knowledge and referent-prevalence and relate it to human elicited endorsements of generalizations.

>>>>

Generalizations are used to make predictions about events or properties of instances that an agent has yet to experience [@HumeTHN].
People readily predict that the next dog will have four legs, drinking another cup of coffee will cause jitters, and a new day will find people that we know doing what they habitually do. 
In each case, we assign a specific exemplar $x$ to a category $k$, and make a prediction that it will have feature $f$.
Probability is the universal currency of belief and this prediction can be described by a conditional probability: $P(x \in f \mid x \in k)$, the probability that $x$ will have $f$ given that it is in $k$ (formally, an instance of $k$ will be in the set of things that have $f$), which we will refer to as the *prevalence* and for convenience write as $p = P(f \mid k)$.
Our *prevalence* $p$ is a *prevalence in the mind*, a latent sense that a future instance of a category would have a particular property; others might call this projectibility [@Goodman1955].
The targets of our predictions can vary widely: They may be objects (a dog), events (a coffee drink), or more fine-grained types (a person on a particular day).
The properties also vary (e.g., having fur, causing jitters, going to the gym).
Yet the mathematical description of the inductive belief is always given by a probability $P(f \mid k)$.

<!-- a useful representation  -->
<!-- How do we come to know things?  -->
<!-- How does one acquire inductive beliefs? -->
Probability is a useful representation for human generalization from observations [@Shepard1987].
If you *observe* several wugs (a novel category) that have two legs, you might infer that all wugs have two legs.
But not all properties have such strong projectibility.
Seeing a wug with broken wings tells you comparatively less about other wugs having broken wings [@Nisbett1983].
Abstract, potentially domain-specific, beliefs about the projectibility of different properties can be represented by a probability distribution over the prevalence $P(p)$ [@Kemp2008].
Bayes Rule provides the way to update one's prior beliefs with some obesrvational data $d$: $P(p \mid d) \propto P(d \mid p) \cdot P(p)$, assuming some generative process that resulted in the observed data---summarized in the likelihood $P(d \mid p)$ [@Tenenbaum2001].

@@

The language of generalization is easy to spot when a property, that could apply to an individual, is predicated of a category (e.g., "Dogs have four legs"; *has four legs* could apply to an individual as in "This dog has four legs").
In the semantics literature on generics, bare plural sentences of this kind are sometimes described as *characterizing sentences* in contrast with *kind-denoting* sentences where the property can only meaningfully apply to the category as a whole (e.g., "Dinosaurs are extinct"; *extinct* cannot apply to an individual dinosaur).
<!-- Dinosaurs are exctinct ~ John is a smoker? -->
Generalization can also manifest when describing instances of an individual [*habitual language*; e.g., "John smokes"; @Carlson1977, @Carlson2005].
Verbs like *causes* or *makes* also seem to convey generalization, in this case, about an instance of an actual causal event (*causal language*; e.g., "Staring at the sun makes you go blind"). 
Psychologists, linguists, and philosophers have long studied the language of generalization, as it appears very simple (e.g., syntactically) yet its meaning is difficult to formalize.^[
  We further distinguish the problem of formalizing a *meaning* for the language of generalization as opposed to the problem of *identifying* generalizations. 
  Syntax alone is neither necessary nor sufficient for a listener to know that the sentence conveys a generalization  (e.g., indefinite singulars can encode generalizations: "A dog has four legs" and bare plurals may not: "Dogs are on my front lawn").
  Our analysis thus begins once a sentence has been disambiguated as conveying a generalization. 
]

If generalization from observations can be described by a probability $p$, it is natural to posit that same construct will be at the heart of a semantic theory of the language of generalization.
In semantics, belief updating generally passes through Boolean *truth values* [@Montague1973].
The simplest way to derive a Boolean from a scalar quantity like probability is via a *threshold semantics*: The utterance is true if the relevant scalar value is above a threshold.
Thus, the simplest semantics for a generalization is a threshold on the prevalence: $\denote{gen}(p, \theta) = \{p >\theta\}$.

This formalism builds on a tradition in formal semantics to try to precisely define the conditions under which a sentence is true or false [@Montague1973].
We go beyond previous semantic models of generics [e.g., @Cohen1999] by embedding this semantic hypothesis in a cognitive model in which relevant contextual knowledge can be formalized.
By doing so, we thus provide an account of how context influences generic understanding and are able to make precise, quantitative predictions about human judgments. /ndg{this paragraph belongs later, after the model has been finished.}

@@

We formalize this underspecified (or, *vague*) semantics in a Bayesian cognitive model, where the threshold $\theta$ is sampled from a uniform prior distribution over possible thresholds $P(\theta)$ [@Lassiter2013;   @Lassiter2015; cf., @Qing2014].^[
  Following @Lassiter2015, we defined $P(\theta)$ to be a uniform distribution over the support of $P(p)$.
]
The model also has prior knowledge about the property $P(p)$ (described in more detail below) and literal interpretation is given by updating these prior beliefs according the truth-functional meaning of the utterance $\delta_{\denote{u}(p, \theta)}$:

@@

Given the above model of literal interpretation, how do we decide when a statement conveying a generalization is intuitively true or not?
We consider the notion of felicity or endorsement from the information-theoretic perspective: Would a generalization statement convey the appropriate information to a naive interpreter? 
We formalize this by considering how well the generalization would convey the target prevalence $p$ to the $Lit$ model (Eq. \ref{eq:L0}).^[
  A more general version of this model can relax the assumption that the endorsement model has access to a specific prevalence $p$ that it wants to communicate.
  Rather, the endorser may have a distribution over $p$ for the category $k$, corresponding to the endorser's uncertain beliefs about the prevalence. 
  In this situation, we would define the endorsement model decision to be with respect to the expected value of the informativity, which integrates over the endorsement model's belief distribution:
  $S(u \mid k) \propto \exp{(\lambda \cdot {\mathbb E}_{p\sim P_{k}} \ln{ \int_{\theta} Lit(p, \theta \mid u)} \diff \theta )}$
  For the empirical case studies described below, these two versions of the model make almost identical predictions.
]

\begin{equation} 
S(u \mid p) \propto \exp{(\lambda \cdot \ln{ \int_{\theta} Lit(p, \theta \mid u)  \diff \theta } )}
\label{eq:S1}
\end{equation}
\ndg{this eqn isn't well motivated. if we never use cost in this paper, then get rid of exp and ln and just have Lit to a power of lambda. also say why it's called S earlier?}

The endorsement model simulates what the interpretation model would believe upon hearing an utterance $u$: $Lit(p, \theta \mid u)$.
$S$ does not know $\theta$ and integrates over the interpreter's posterior beliefs about $\theta$.
Following Bayesian decision theory, we assume the decision to endorse is approximately rational, with degree of rationality given by $\lambda$. \ndg{this doesn't make a ton of sense in this context, since we haven't set up S as a decision maker....}
The proportionality in Eq. \ref{eq:S1} implies normalization over a set of alternative utterances. 
We interpret the endorsement task (e.g., "true" vs "false"; "agree" vs. "disagree") as supplying two alternatives: *endorse* or *not*, where *not* we take to be a silent utterance that conveys no information content [@Franke2014cogsci; @Degen2014]: $\denote{null}(x, \theta) = \text{True}$.^[
  This "null" alternative to the generalization can be realized in at least two other ways: the negation of the generalization (e.g., "It is not the case that Robins lay eggs") or the negative generalization (i.e., "Robins do not lay eggs").
  The latter is different because it creates a new threshold (analogous to how "short" and "tall" are modeled with different thresholds).
  All results reported are similar for these two alternatives, and we use the alternative of the "silent" utterance for simplicity.
]

Our endorsement model (Eq. \ref{eq:S1}) provides a mapping from the prevalence in the category being referred to (i.e., how many robins lay eggs)---the *referent prevalence* $p$---to an endorsement probability for the corresponding generalization (e.g., "Robins lay eggs"): $S(u = u_{gen} \mid p)$. 
The model assumes a threshold semantics based on probability (prevalence) but does not require specifying the threshold *a priori* (note that there is no $\theta$ in the left-hand side of Eq. \ref{eq:S1}).
This model can be seen as a special case of a Rational Speech Act model [@Frank2012; @Goodman2016] with only a single level of recursion and the space of possible utterances maximally constrained; the endorsement model is a "speaker", while the interpretation model is a "listener", and we are interested in whether a reasonable speaker would bother to utter the generalization. \ndg{i wonder if the content of this paragrph should be earlier, to motivate the endorsement model?}

@@

The uncertain, truth-conditional threshold $\theta$ is resolved by the literal interpretation model $Lit$ (Eq. \ref{eq:L0}) by integrating the model's prior beliefs about the feature in question---the *prevalence prior* $P(p)$---with the underspecified semantics.
Because $P(\theta)$ is assumed to be a uniform distribution across all contexts, a generalization could in principle be felicitous with any referent-prevalence. 
However, the interpretation model's prior beliefs $P(p)$ will be constrain the interpretation and thus the endorsement.

@@

\ndg{this paragraph should get moved elswhere. it is distracting here....}
Beliefs about properties result from internal, conceptual models of the world. 
We do not take up the task of formalizing particular conceptual models, but rather merely try to measure their implications on beliefs about prevalence. 
Without appealing to domain specific knowledge, one can construct a prior distributions over prevalence by considering the prevalence of the feature among alternative categories.
For example, many different kinds of animals *do not bark* (more precisely: have close to 0 exemplars who bark); thus, the prevalence prior for the feature *barks* should have substantial probability mass near 0, because one would expect a randomly-sampled category to have 0 or close to 0 exemplars that bark.
Furthermore, among the categories that do have exemplars which bark (e.g., *dogs*, *seals*, ...), the probability is quite high that an exemplar will bark; thus, the corresponding prevalence prior would have some probability mass at relatively high prevalences, potentially with some uncertainty (Figure\ \@ref(fig:simulations); column 1).
Abstract beliefs between different kinds of categories could manifest in these prevalence priors in the form of multi-modal distributions; for example, the prevalence prior for *lays eggs* should similarly have substantial probability mass near 0 (since many animal categories do not have egg-layers) but should have a secondary component peaked around 0.5, because among the animals with egg-layers, only the female members of the category have the property (Figure\ \@ref(fig:simulations); column 4).

@@

```{r uniformPriorSimulation, fig.cap = "Computational model behavior. A: Interpretation model posteriors assuming different fixed thresholds. High thresholds rule out more world-states (prevalence levels), leaving fewer world-states among which to distribute the full probability mass. B: Full interpretation model marginalizes over possible thresholds to return a posterior distribution that favors higher prevalence levels in a graded manner. C: Endorsement model predicts higher rates of endorsemens as prevalence levels increase. \ndg{say what P(p) is used here...}", fig.width = 13, fig.asp = 0.3, out.width = "100%", fig.align = "center", cache = F}

@@

These simulations show how the model predicted endorsement depends upon the background knowledge of the interpreter about the property $P(p)$ as well as the referent-prevalence $p$ presented to the endorser. 
We chose schematic values for each of these, and saw that the endorsement model recovers standard intuitions about the truth or falsity of classic examples from the generics literature.
In what follows, we test this theory for a wide range of generalizations, including generalizations of different types (categories, events, and causes).
We do this by both measuring and manipulating background knowledge and referent-prevalence and relate it to human elicited endorsements of generalizations.
